Estimating the number of clusters using Cross Validation

### 2015-06-23 cvclust.tex
finish revising self-consistency proof

\begin{theorem}\label{thm:self-consistency}
Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
Suppose that
\begin{enumerate}[label=(\roman*)]

  \item There are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with the $g$th
    cluster center partitioned as $\mu(g) = \bigl(\muX(g), \muY(g)\bigr)$ with
    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, K$.

  \item The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct; likewise,
    $\muY(1), \dotsc,\muY(K)$ are all distinct.

  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.

  \item The training set contains at least one member of each cluster:
    for all $g$ in the range $1, \dotsc, K$, there exists at least one $i$ in
    the range $1, \dotsc, n$ such that $G_i = g$.

@@

\begin{proposition}
If assumptions $1, 2, 3, 5$ hold, then there exists a classification method
works perfectly in case of ($k \leq K$): the re-substitution error is $0$.
That is, for $\forall x_i \in \mathbf{X}_{train}$ prediction on $x^L_i$ always
correctly predict cluster where $x^R_i$ is in.
\end{proposition}
\begin{proof}
To train classifier $\mathcal{C}$, we have training set
$\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \{\left(x^L_i,\hat{g}
\left(x^R_i\right)\right), i =1,2,...,m\}$. By assumption $5$, $\forall i$
$x_i = \mu_j$, $j=1,2,...,K$ and $k \leq K$. Hence the training set has $K$
unique pairs $\{\left(\mu^L_j, \hat{g}
\left(\mu^R_j\right)\right),j=1,2,...,K\}$, with $K$ unique $\mu^L_j$ by
assumption $3$ and $k$ unique $\hat{g} \left(\mu^R_j\right)$. This is because
each pair $(\mu^L_j,\mu^R_j)$ is unique by assumption $2$ and
$\left(\mu^R_j,\hat{g} \left(\mu^R_j\right)\right)$ is unique since K-means
assigns one and only one label to each $\mu^R_j$. Define the classifier
$\mathcal{C}$ as following: \[ \mathcal{C}(x^L_i) = \sum^K_{j=1}\hat{g}
\left(\mu^R_j\right) \cdot I_{\{c_i = j\}}  \] It's easy to see such
classifier has re-substitution error $0$.
\end{proof}

The proof of above theorem follows from following two lemmas:
\begin{lemma}
Suppose that the assumptions of Theorem $1$ are in force.  If $k < K$, then
$\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all $K$ distinct values
in $\{ \muY(1), \dotsc, \muY(K) \}$.  However, there are only $k$ distinct
values of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}
	If the assumptions of Theorem $1$ hold, when $k=K$, the CV Error $=0$
\end{lemma}
\begin{proof}
	It's sufficient to show that when $k=K$, $Error_k=0$. By assumption $3$ and $5$, training set contains $K$ centers. Which also implies $\mathbf{X}^R_{train}$ contains $K$ distinct $x^R_i$ $(\mu^R_1,...,\mu^R_K)$. When apply K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the unique solution of 
	\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
	is 
\begin{equation}	
	\{\hat{\mu}^R \left(g_j\right) = \mu^R_j, j=1,...,K\}
\end{equation}	
with 
\begin{equation}	 
	\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}	
Where $\hat{g} \left(x^R_i\right) \in  \mathcal{G}$ is the assigned cluster of $x^R_i$ by K-means. One can easily check that such solution is unique and above objective function attains its lower bound $0$.\\\\
By Proposition $1$, $\mathcal{C}$ has re-substitution error $0$, which means its prediction error is $0$ on $\mathbf{X}^L_{test}$ as $\forall x^L_i \in \mathbf{X}^L_{test}, x^L_i \in \mathbf{X}^L_{train}$ by assumption $1,3,5$. Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$. By equation $(3)$, we have $x^R_i = \mu^R \left(c_i\right) = \mu^R_j$ if $\hat{g} \left(x^L_i\right) = \hat{g} \left(x^R_i\right) = g_j$. Therefore,
\[ Error_K = \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 =  \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - \mu^R_j||^2= 0\]
The last equation holds because of equation $(2)$.
\end{proof}

>>>>

\begin{theorem}\label{thm:self-consistency}

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.

  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.

  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.

  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.

@@

This theorem is implied by the following two lemmas.

\begin{lemma}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k < K$, then $\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
$\{ \muY(1), \dotsc, \muY(K) \}$.
Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
exactly $K$ such distinct values.  However, there are only $k$ distinct values
of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k \geq K$, then $\CV(k) = 0$.
\end{lemma}
\begin{proof}
From assumptions~\ref{asn:self-consistency-noiseless},
\ref{asn:self-consistency-distinct-muy},
and~\ref{asn:self-consistency-train}, we know the cluster centers
gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
assumption~\ref{asn:self-consistency-noiseless} again, we get that
$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.

Since assumption~\ref{asn:self-consistency-distinct-mux} ensures that
$\muX(1), \dotsc, \muX(K)$ are all distinct, we must have that $\hGX_i = G_i$
for all $i = 1, \dotsc, m+n$.  In particular, this implies that $\bmuY(\hGX_i)
= Y_i$ for $i = 1, \dotsc, m+n$, so that $\CV(k) = 0$.
\end{proof}
